\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-22 paper/paper.tex
ref memo start
simulations with age and change Hahner color

With respect to our second measure of marathon skill, Figure
\ref{fig:45degreeplot_half} shows that there was nothing abnormal
about the Hahner twins' overall finishing times, conditional on their
half marathon splits. As one might expect, a runner's time halfway
through the course is a fairly strong predictor of her finishing time.
Here we see that the relationship between the Hehner twin's half
marathon split times and their finishing times is similar to that of
other Rio runners.  The two points representing the Hahners are in the
middle of the distribution of points and therefore imply that the
twins do not seem to have slowed down in any unusual way.  This is
consistent with the previous Figure \ref{fig:45degreeplot} and
inconsistent with accusations made against the Hahner twins, at least
the part of the accusation that focused on their overall pace.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's orange dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

@@

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to subsequent DNFs.  The usual color scheme applies here,
and the Estonian triplet DNF prior to the half marathon split is
evident. As noted in the figure, Lisa Hahner was ahead of her sister
through 15 kilometers at which point Anna took the lead.  Figure
\ref{fig:secondsbehind} contains two red dots representing the North
Korean Kim twins, but this is not visually apparent because the Kim
twins had identical split times during the entire marathon.

@@

To estimate the conditional distribution of each Rio final results, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of marathon skill $X_i$. Here
we start by assuming that marathon skill can be measured by personal
best marathon time:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.  We
estimate $\beta_1$, $\beta_2$, and $\sigma$ using ordinary least
squares and all finishing Rio marathoners except for the Hahner/Kim
twins and Luik triplets.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anne and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her best time squared.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on personal best times}
  \label{fig:simdiff}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on half marathon splits}
  \label{fig:simdiffhalf} 
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_half.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

>>>>

With respect to our second measure of marathon skill, Figure
\ref{fig:45degreeplot_half} shows that there was nothing abnormal
about the Hahner twins' overall finishing times, conditional on their
half marathon splits. As one might expect, a runner's time halfway
through the course is a fairly strong predictor of her finishing time.
Here we see that the relationship between the Hahner twin's half
marathon split times and their finishing times is similar to that of
other Rio runners.  The two points representing the Hahners are in the
middle of the distribution of points and therefore imply that the
twins do not seem to have slowed down in any unusual way.  This is
consistent with the previous Figure \ref{fig:45degreeplot} and
inconsistent with accusations made against the Hahner twins, at least
the part of the accusation that focused on their overall pace.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's red dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

@@

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to subsequent DNFs.  The usual color scheme applies here,
and the Estonian triplet DNF prior to the half marathon split is
evident. As noted in the figure, Lisa Hahner was ahead of her sister
through 15 kilometers at which point Anna took the lead.  Figure
\ref{fig:secondsbehind} contains two orange dots representing the North
Korean Kim twins, but this is not visually apparent because the Kim
twins had identical split times during the entire marathon.

@@

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of the runner's marathon skill $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + \beta_3\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution. Taking each runner's 
personal best marathon time as a proxy for their marathon skill, we
estimate $\beta_1$, $\beta_2$, $\beta_3$ and $\sigma$ using ordinary least
squares on all finishing Rio marathoners.  We exclude the Hahner/Kim
twins and Luik triplets from the sample so that that our estimates are not affected by any intentional twin/triplet finish.  
For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$, her best time squared $X^2_{i}$, and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and $\tilde{\beta}_3$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and
  $\tilde{\beta}_3$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{\beta}_3\,Z_i +  \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on personal best times}
  \label{fig:simdiff}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_with_age.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_with_age.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on half marathon splits}
  \label{fig:simdiffhalf} 
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_half_with_age.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes_half}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half_with_age.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks_half}
  \end{subfigure}
\end{figure}
